   %% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Preface}
This thesis consists of an introduction, materials and methods, three research chapters, and discussion followed by bibliography.

All animal experiments were conducted according to the Canadian Council on Animal Care guidelines, approved by the Animal Care Committee of the University of British Columbia, according to the guidelines of the Institutional Animal Care and Use Committee (IACUC) under protocols A14-0290 and A15-0248.
%------------------------------------------------------------------------
  
  \textbf{(i) Chapter 3:}

All experiments in this chapter were designed, conducted and analysed by me, unless stated otherwise, with intellectual inputs from Dr. Sam Aparicio. PDX tumors palpation were done by animal technicians (Aparicio lab), Ms. Teresa Ruiz de Algara and Ms. So Ra Lee. They also helped in initial transplants into \ac{NSG} mice but all re-transplants into \ac{NRG} mice were done by me. 


From the drug efficacy \textit{in vitro} cultures, a few experiments contributed to the publication of \textit{Nature Communication, 2017} paper \cite{xu2017cx}:

\textbf{CX-5461 is a DNA G-quadruplex stabilizer with selective lethality in BRCA1/2 deficient tumours.}

Xu H, Di Antonio M, McKinney S, Mathew V, Ho B,O'Neil NJ,Santos ND,Silvester J, Wei V, Garcia J, \emph{\textbf{Kabeer F}}, Lai D, Soriano P, Banath J, Chiu DS, Yap D, Le DD, Ye FB, Zhang A, Thu K, Soong J, Lin SC, Tsai AH, Osako T, Algara T, Saunders DN, Wong J, Xian J, Bally MB, Brenton JD, Brown GW, Shah SP, Cescon D, Mak TW, Caldas C, Stirling PC, Hieter P, Balasubramanian S, Aparicio S. Nat Commun. 2017 Feb 17;8:14432, PMID: 28211448.



 \textit{In vivo} drug sensitivity experiments were all done by me including, tumor transplantation, drug dosing, monitoring, tumor measurement and tumor growth records. I generated the growth curve graphs in consultation with a statistician, Dr. Steven McKinney.


The other half of chapter 3 is a \textit{Genome biology} paper \cite{o2019dissociation}:


\textbf{Dissociation of solid tumor tissues with cold active protease for single-cell RNA-seq minimizes conserved collagenase-associated stress responses.}

O'Flanagan CH*, Campbell KR*, Zhang AW*, \emph{\textbf{Kabeer F*}}, Lim JLP, Biele J, Eirew P, Lai D, McPherson A, Kong E,
Bates C, Borkowski K, Wiens M, Hewitson B, Hopkins J, Pham J, 
Ceglia N, Moore R, Mungall AJ,
McAlpine JN; CRUK IMAXT Grand Challenge Team, Shah SP, Aparicio S.

\textbf{*equal contribution.} Genome Biol. 2019 Oct 17;20(1):210, PMID: 31623682 

All of the PDX experiments were designed, executed, harvested by myself, dissociations in collaboration with Dr. Ciara O'Flanagan, statistical methods and analysis Dr. Kieran Campbell (statistician).

The established PDX tumors and dissociation optimizations done for chapter 3, also helped in 3 more publications from Aparicio lab, including \cite{laks2019clonal, campbell2019clonealign, de2020epiclomal}:
     
\textbf{Clonal Decomposition and DNA Replication States Defined by Scaled Single-Cell Genome Sequencing.}
Laks E, McPherson A, Zahn H, Lai D, Steif A, Brimhall J, Biele J, Wang B, Masud T, Ting J, Grewal D, Nielsen C,Leung S,Bojilova V,Smith M, Golovko O, Poon S, Eirew P, \emph{\textbf{Kabeer F}}, Ruiz de Algara T, Lee SR, Taghiyar MJ, Huebner C, Ngo J, Chan T, Vatrt-Watts S, Walters P, Abrar N, Chan S, Wiens M, Martin L, Scott RW, Underhill TM, Chavez E, Steidl C, Da Costa D, Ma Y, Coope RJN, Corbett R, Pleasance S, Moore R, Mungall AJ, Mar C, Cafferty F, Gelmon K, Chia S; CRUK IMAXT Grand Challenge Team, Marra MA, Hansen C, Shah SP, Aparicio S. Cell. 2019 Nov 14;179(5):1207-1221.e22,  PMID: 31730858.


\textbf{clonealign: statistical integration of independent single-cell RNA and DNA sequencing data from human cancers.}
Campbell KR, Steif A, Laks E, Zahn H, Lai D, McPherson A, Farahani H, \emph{\textbf{Kabeer F}}, O'Flanagan  C, Biele J, Brimhall J, Wang B, Walters P; IMAXT Consortium, Bouchard-Cote A, Aparicio S, Shah SP. Genome Biol. 2019 Mar 12;20(1), PMID: 30866997. 

\textbf{Epiclomal: Probabilistic clustering of sparse single-cell DNA methylation data.}
P E de Souza C, Andronescu M, Masud T, \emph{\textbf{Kabeer F}}, Biele J, Laks E, Lai D, Ye P, Brimhall J, Wang B, Su E, Hui T, Cao Q, Wong M, Moksa M, Moore RA, Hirst M, Aparicio S, Shah SP. PLoS Comput Biol. 2020 Sep 23;16(9):e1008270. doi: 10.1371/journal.pcbi.1008270. Online. PMID: 32966276

%-------------------------------

  \textbf{(ii) Chapter 4:}

This chapter is based on a manuscript that is currently in press at \textit{Nature, 2021}.


 \textbf{Single  cell  fitness  landscapes  induced  by  p53  deletion and platinum perturbation in cancer}

Sohrab Salehi*, \emph{\textbf{Farhia Kabeer*}}, Nick Ceglia, Mirela Andronescu, Marc Williams, Kieran R. Campbell, Tehmina Masud, Beixi Wang, Justina Biele, Jazmine Brimhall, Jerome Ting, Allen Zhang, Ciara O'Flanagan , Fatemeh Dorri, Nicole Rusk, Emma Laks, Hakwoo Lee, Teresa Algara, Sora Lee, Brian Yu Chieh Cheng, Peter Eirew, Takako Kono, Jennifer Pham, Diljot Grewal, Daniel Lai, Richard Moore, Andrew J. Mungall, Marco A Marra, IMAXT Consortium, Andrew McPherson, Alexandre Bouchard-Cote, Samuel Aparicio+, Sohrab P. Shah+

 \textbf{*equal contribution.} In press at \textit{Nature}, 2021

https://www.biorxiv.org/content/10.1101/2020.05.08.081349v1.full.pdf
\\
 All PDX experiments were designed and executed by me over several years of the project, from tumor serial transplants, drug dosing, monitoring, tumor collection and tumor dissociations. Single cell sequencing from tissues, I obtained and prepared, was executed by the single cell sequencing core at the BCCRC. I conducted interpretation of the data, designed figures and wrote the manuscript jointly with another PhD student, Sohrab Salehi, who developed the Wright-Fisher model and the Bayesian phylogenetic inference method \texttt{sitka}.  

%------------------------------------------------------------------

  \textbf{(iii) Chapter 5:}
   
This chapter is based on a manuscript in preparation. I designed the experiments, performed the transplants and drug treatments, harvested tumours and prepared cells for submission to the single cell sequencing core. I designed the figures and interpreted the data with support from Dr. Tran and Dr. Andronescu, statisticians, who assisted with high dimensional data reduction and expression network analysis.
  
\textbf{Longitudinal tracking of drug-induced transcriptomic reprogramming in triple negative breast cancer pre-clinical model} \textit{(Manuscript in preparation)}

\emph{\textbf{Farhia Kabeer*}}, Hoa Tran*, Mirela Andronescu*,   
Nicholas Ceglia, Hakwoo Lee, Sohrab Salehi, Beixi Wang, Justina Biele, Jazmine Brimhall, David Gee, Steven Mckinney, Ciara O'Flanagan, Teresa Algara, Takako Kono, Jennifer Pham, Daniel Lai, Richard Moore, Andrew J. Mungall, IMAXT Consortium,  Andrew  McPherson, Andrew Roth, Kieran R. Campbell, Sohrab P. Shah, Samuel Aparicio.